From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-03-06 appendixA/appendixA.Rmd
tweaking refs

```{r include=FALSE}
## note: needs library Cairo
knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="cairo_pdf")
```

```{r libraries, cache=FALSE, message=FALSE}
library(nimble)
library(RcppRoll)
## devtools::install_github("cboettig/regimeshifts")
library(regimeshifts) 
library(adaptivetau)  ## For Gillespie simulation only
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(scales)
## Set a default color pallete
colours <- ptol_pal()(2)
```

```{r echo=FALSE, cache=FALSE, eval=TRUE, message=FALSE}
## plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
library(Cairo)
library(hrbrthemes)
library(extrafont)

## One-time setup
hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
#d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
#d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
#write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)

## load fonts (needed each time)
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())

@@

e <- 0.2
c <- 1

## Run simulations
sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
                       transitions, rateF,
                       params = list(c=c, e=e, N=100),
                       tf=30) %>% as_tibble()

@@

We will also create a separate data frame (`tibble`) corresponding to our theoretical predictions from the system size expansion.  We use popular functions from the `tidyverse` suite of R packages for the manipulation of `data.frames`.  Below, the `mutate` function is used to add new columns for plus and minuse one standard deviation based on theoretical calculation of $\sqrt{N\tfrac{e}{c}}$, and the `map_dfr` function is used to apply this to both `small` and `large` cases.  We use `tidyverse` syntax  [see @Ross2017] instead of base R in these examples for the following reasons:

@@

```{r}
## combine data.frames, using column called "system_size"
## to indicate large noise or small noise simulations
gillespie <- 
  bind_rows("large" = sim_large, 
            "small" = sim_small,  
            .id = "system_size") %>% 
  bind_rows(theory)

## Write data to file
write_csv(gillespie, "gillespie.csv")
```

Example plotting command to generate the figure as shown in main text:

@@

Simulations of the quasi-cycles can easily be performed in NIMBLE, which allows us to express the stochastic models in the familiar BUGS syntax.  Change the values of the model and noise parameters below to explore these results.  While the models below can easily be implemented in base R, the BUGS syntax for expressing a stochastic model provides a more generic formulation that can easily be used in parameter estimation as well; see @nimble.  This can also imporve computational performance because @nimble compiles these models into C++ code.

```{r}
## Define the BUGS model:
quasicycle  <- nimble::nimbleCode({
  x[1] <- x0
  y[1] <- y0
  for(t in 1:(N-1)){
    mu_x[t] <- x[t] + x[t] * r * (1 - x[t] / K) - b * x[t] * y[t]
    x[t+1] ~ dnorm(mu_x[t], sd = sigma_x)
    mu_y[t] <- y[t] + c * x[t] * y[t] - d * y[t]
    y[t+1] ~ dnorm(mu_y[t], sd = sigma_y)
  }
})

## creates a data.frame of parameter combinations for each noise level
p <- 
  data.frame(
  data.frame(r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800),
  data.frame(sigma_x = c(.00001,0.01),  sigma_y = c(.00001,0.01)))

## define a function to perform a simulation given a set of parameters (constants)
f <- function(constants){
  model <- compileNimble(nimbleModel(quasicycle, 
                                     constants = constants, 
                                     inits = list(x0 = 1, y0 = 1)))
  ## set random seed for reproducibility
  set.seed(123)
  ## Run the actual simulation, using nimble
  simulate(model)

  ## Assemble the results into a data.frame (tibble).  
  tibble(t = seq_along(model$x), 
         x = model$x, 
         y = model$y, 
         sigma = constants$sigma_x)
}

## Do this simulation for each set of parameters, given 
quasicycle_df <- p %>% rowwise() %>% do(f(.))

## Write results to file for later reference and plotting
write_csv(quasicycle_df, "quasicycles.csv")
```

From this quasicycle simulation data we can recreate the plot shown in Figure 2 as follows:

```{r quasicycles, fig.width=7}
read_csv("quasicycles.csv") %>%
  ## Data in long from
  gather(species, population, -t, -sigma) %>%
  ## Re-arrange and and re-lable for the plots:
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
```

@@

Our example of a stochastic osccilato is based on May's model:

@@

may  <- nimble::nimbleCode({

  x[1] <- x0
  for(t in 1:(N-1)){
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ Q / (x[t] ^ Q + H ^ Q)
    y[t+1] ~ dnorm(mu[t], sd = sigma)
    x[t+1] <- max(y[t+1],0)
  }

@@

Seperately, we will also compute the growth and consumption curves and the potential function for comparison with theory:

```{r}
theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  ## As before, we add columns with mutate
  mutate(g = x * p$r * (1 - x / p$K), 
         c = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  mutate(potential = - cumsum(g - c)) %>%
  ## assemble resulting data into "long" form with the
  ## column headings as named here:
  gather(curve, y, -x, -potential)

@@

```{r}
max_days <- 1000
## Perform three simultaneous simulations, x,y,z, using
## different dilution factors DF

## Intialize variables
z <- y <- x <- numeric(max_days)
z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5

## set random seed for reproduciblity
set.seed(123)

## simulate for max_days
for(day in 1:(max_days-1)){
  x[day+1] <- dai(x[day], epsilon = rnorm(1,0, 0.001), DF = 1790)
  y[day+1] <- dai(y[day], epsilon = rnorm(1,0, 0.001), DF=1799)
  z[day+1] <- dai(z[day], epsilon = rnorm(1,0, 0.001), DF = 1800)

@@

## gather results into long data.frame 
df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
  gather(series, pop, -t)

## and plot results
ggplot(df, aes(t, pop, col=series)) + 
  geom_line()

@@

We will also need a function for rolling autocorrelation for our warning sign calculation:

@@

## Simulation of Dai model:

```{r}
## Slow environmental degredation through small stepwise changes
## (increasing the serial dilution factor, as in the Dai et al experiment.)
DF <- as.numeric(sapply(seq(0, 2000, length=9), rep, 40))

@@

## also compute warning signs, include in single table
tip_df %>% 
  mutate(autocorrelation = roll_acor(x, n = 100, fill=NA),
         variance = roll_var(x, n = 100, fill=NA),
         tip_time = tip_time) %>%
  write_csv("tipping.csv")
```

We are now ready to create the panels shown in Figure 4:

@@

p1 <- tipping %>%
  select(-x) %>%
  filter(t < 1866) %>%
  rename("A. autocorrelation" = autocorrelation, "variance" = variance) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line(col=colours[[1]]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y") +
  scale_y_continuous(labels=scientific)
p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + 
  geom_line(col=colours[1]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) +
  scale_y_log10(labels=scientific)  + 
  labs(subtitle = "C. Yeast population size", 
       caption = "Vertical red dashed line indicates tipping point location")

@@

Note that @Dai2015 plots coefficient of variation, (standard deviation over mean), rather than the variance.  This is a popular choice since the resulting indicator is a dimensionless measure of the noise relative to the mean, but also confounds changes in the mean dynamics with changes in the noise dynamics.  This can give a misleading impression of the performance of the indicator: in many models, the mean (that is, the value of the equilibrium population size, $\hat x$) will also decrease prior to the transition, resulting in a more pronounced increase in the coefficent of variation than in the variance.  However, changes in mean per se should are not a generic indicator of stability predicted by the theory -- indeed, the whole point of the theory was to detect changes that would not be manifested in the mean.  In general, systems going through a saddle-node bifurcation can vary greatly in how much the mean changes.  The mean will generally decrease, since it must eventually meet the 'saddle' point below it -- it is a question of whether the saddle point does most of the increasing or the node does most of the decreasing.

@@

Analytical predicted population size, as a fraction of $K$, is $\tfrac{1}{2} + \tfrac{1}{2} \sqrt{1- 8\sigma^2/K^2}$.  Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.  Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  As predicted by the system size expansion, the deviation from the deterministic result is of order $N^{-1}$.

@@

## Parameters
p <- list(r = .4, K = 1, sigma = .15, N = 500)
equib <- (1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2

## BUGS model
logistic  <- nimble::nimbleCode({
  x[1] <- x0
  for(t in 1:(N-1)){
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
    x[t+1] ~ dnorm(mu[t], sd = sigma)
  }
})

## Compile and run model
model <- 
nimbleModel(logistic,
            constants = p,
            inits = list(x0 = 1))
cmodel <- compileNimble(model)
set.seed(123)

## Perform the simulations
df <- map_dfr(1:10000, 
              function(rep){
                simulate(cmodel)
                data.frame(t = seq_along(cmodel$x), x = cmodel$x)
              },
              .id = "rep")

@@

```{r}
read_csv("inflation.csv") %>% 
  gather(statistic, value, -t) %>%
  ggplot(aes(t,value, col=statistic, lty=statistic)) + 
  geom_line(lwd=1) + 
  scale_color_ptol() + 
  coord_cartesian(xlim=c(100,500), ylim=c(1.2, .5))
```

>>>>

```{r include=FALSE}
# note: needs library Cairo
knitr::opts_chunk$set(message=FALSE, cache = 1, warning = FALSE,  dev="cairo_pdf")
```

```{r libraries, cache=FALSE, message=FALSE}
library(nimble)
library(RcppRoll)
# devtools::install_github("cboettig/regimeshifts")
library(regimeshifts) 
library(adaptivetau)  # For Gillespie simulation only
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(scales)
# Set a default color pallete
colours <- ptol_pal()(2)
```

```{r echo=FALSE, cache=FALSE, eval=TRUE, message=FALSE}
# plot themes, non-essential. Comment out if theme creates trouble with fonts on your system.
library(Cairo)
library(hrbrthemes)
library(extrafont)

# One-time setup
hrbrthemes::import_roboto_condensed()   # One-time use to set up fonts 
#d <- read.csv(extrafont:::fonttable_file(), stringsAsFactors = FALSE)
#d[grepl("Light", d$FontName), ]$FamilyName <- "Roboto Condensed Light"
#write.csv(d, extrafont:::fonttable_file(), row.names = FALSE)

# load fonts (needed each time)
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())

@@

e <- 0.2 # patch death rate
c <- 1   # patch colonization rate

# Run simulation for small and large N
sim_small <- adaptivetau::ssa.exact(init.values = c(n = 50),
                       transitions, rateF,
                       params = list(c=c, e=e, N=100),
                       tf=30) %>% as_tibble()

@@

We will also create a separate data frame (`tibble`) corresponding to our theoretical predictions from the system size expansion.  We use popular functions from the `tidyverse` suite of R packages for the manipulation of `data.frames`.  Below, the `mutate` function is used to add new columns for plus and minus one standard deviation based on theoretical calculation of $\sqrt{N\tfrac{e}{c}}$, and the `map_dfr` function is used to apply this to both `small` and `large` cases.  We use `tidyverse` syntax  [see @Ross2017] instead of base R in these examples for the following reasons:

@@

```{r}
# combine data.frames, using column called "system_size"
# to indicate large noise or small noise simulations
gillespie <- 
  bind_rows("large" = sim_large, 
            "small" = sim_small,  
            .id = "system_size") %>% 
  bind_rows(theory)

# Write data to file
write_csv(gillespie, "gillespie.csv")
```

Example plotting command to generate the figure as shown in main text, (using `ggplot2` plotting methods from `tidyverse`):

@@

Simulations of the quasi-cycles can easily be performed in NIMBLE, which allows us to express the stochastic models in the familiar BUGS syntax.  Change the values of the model and noise parameters below to explore these results.  While the models below can easily be implemented in base R, the BUGS syntax for expressing a stochastic model provides a more generic formulation that can easily be used in parameter estimation as well; see @deValpine2017 for details.  This can also improve computational performance because NIMBLE compiles these models into C++ code.

```{r}
# Define the BUGS model:
quasicycle  <- nimble::nimbleCode({
  x[1] <- x0
  y[1] <- y0
  for(t in 1:(N-1)){
    # Deterministic terms look like normal R code:
    mu_x[t] <- x[t] + x[t] * r * (1 - x[t] / K) - b * x[t] * y[t]
    # Stochastic assignment uses ~ to indicate 'distributed as' normal (Gaussian)
    x[t+1] ~ dnorm(mu_x[t], sd = sigma_x)
    mu_y[t] <- y[t] + c * x[t] * y[t] - d * y[t]
    y[t+1] ~ dnorm(mu_y[t], sd = sigma_y)
  }
})

# creates a data.frame of parameter combinations for each noise level
p <- 
  data.frame(
  data.frame(r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800),
  data.frame(sigma_x = c(.00001,0.01),  sigma_y = c(.00001,0.01)))

# define a function to perform a simulation given a set of parameters (constants)
f <- function(constants){
  model <- compileNimble(nimbleModel(quasicycle, 
                                     constants = constants, 
                                     inits = list(x0 = 1, y0 = 1)))
  # set random seed for reproducibility
  set.seed(123)
  # Run the actual simulation, using nimble
  simulate(model)

  # Assemble the results into a data.frame (tibble).  
  tibble(t = seq_along(model$x), 
         x = model$x, 
         y = model$y, 
         sigma = constants$sigma_x)
}

# Do this simulation for each set of parameters, given 
quasicycle_df <- p %>% rowwise() %>% do(f(.))

# Write results to file for later reference and plotting
write_csv(quasicycle_df, "quasicycles.csv")
```

From this quasi-cycle simulation data we can recreate the plot shown in Figure 2 as follows:

```{r quasicycles, fig.width=7}
read_csv("quasicycles.csv") %>%
  # Data in long from
  gather(species, population, -t, -sigma) %>%
  # Re-arrange and and re-lable for the plots:
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() 
```

@@

Our example of a stochastic oscillator is based on May's model:

@@

# Define stochastic model in BUGS notation
may  <- nimble::nimbleCode({

  x[1] <- x0
  for(t in 1:(N-1)){
    # Determinstic mean looks like standard R
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ Q / (x[t] ^ Q + H ^ Q)
    # Note the use of ~ in BUGS to show 'distributed as normal' 
    y[t+1] ~ dnorm(mu[t], sd = sigma)
    x[t+1] <- max(y[t+1],0)
  }

@@

Separately, we will also compute the growth and consumption curves and the potential function for comparison with theory:

```{r}
theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  # As before, we add columns with mutate
  mutate(g = x * p$r * (1 - x / p$K), 
         c = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  mutate(potential = - cumsum(g - c)) %>%
  # assemble resulting data into "long" form with the
  # column headings as named here:
  gather(curve, y, -x, -potential)

@@

```{r}
max_days <- 1000
# Perform three simultaneous simulations, x,y,z, using
# different dilution factors DF

# Intialize variables
z <- y <- x <- numeric(max_days)
z[1] <- y[1] <- x[1] <- 7e4 # 1.7e5

# set random seed for reproduciblity
set.seed(123)

# simulate for max_days
for(day in 1:(max_days-1)){
  x[day+1] <- dai(x[day], epsilon = rnorm(1,0, 0.001), DF = 1790)
  y[day+1] <- dai(y[day], epsilon = rnorm(1,0, 0.001), DF=1799)
  z[day+1] <- dai(z[day], epsilon = rnorm(1,0, 0.001), DF = 1800)

@@

# gather results into long data.frame 
df <- data.frame(t = 1:max_days, x = x, y = y, z = z) %>%
  gather(series, pop, -t)

# and plot results
ggplot(df, aes(t, pop, col=series)) + 
  geom_line()

@@

We will also need a function for rolling auto-correlation for our warning sign calculation:

@@

### Simulation of Dai model:

```{r}
# Slow environmental degredation through small stepwise changes
# (increasing the serial dilution factor, as in the Dai et al experiment.)
DF <- as.numeric(sapply(seq(0, 2000, length=9), rep, 40))

@@

# also compute warning signs, include in single table
tip_df %>% 
  mutate(autocorrelation = roll_acor(x, n = 100, fill=NA),
         variance = roll_var(x, n = 100, fill=NA),
         tip_time = tip_time) %>%
  write_csv("tipping.csv")
```

We are now ready to create the panels shown in Figure 4.  A little data tidying and `ggplot` layering creates the final plot:

@@

p1 <- tipping %>%
  select(-x) %>%
  filter(t < 1866) %>%
  rename("A. autocorrelation" = autocorrelation, "variance" = variance) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line(col=colours[[1]]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y") +
  scale_y_continuous(labels=scientific)
p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + 
  geom_line(col=colours[1]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) +
  scale_y_log10(labels=scientific)  + 
  coord_cartesian(ylim=c(1e3, 5e5)) + 
  labs(subtitle = "C. Yeast population size", 
       caption = "Vertical red dashed line indicates tipping point location")

@@

Note that @Dai2015 plots coefficient of variation, (standard deviation over mean), rather than the variance.  This is a popular choice since the resulting indicator is a dimensionless measure of the noise relative to the mean, but also confounds changes in the mean dynamics with changes in the noise dynamics.  This can give a misleading impression of the performance of the indicator: in many models, the mean (that is, the value of the equilibrium population size, $\hat x$) will also decrease prior to the transition, resulting in a more pronounced increase in the coefficient of variation than in the variance.  However, changes in mean per se should are not a generic indicator of stability predicted by the theory -- indeed, the whole point of the theory was to detect changes that would not be manifested in the mean.  In general, systems going through a saddle-node bifurcation can vary greatly in how much the mean changes.  The mean will generally decrease, since it must eventually meet the 'saddle' point below it -- it is a question of whether the saddle point does most of the increasing or the node does most of the decreasing.

@@

Analytically predicted population size, as a fraction of $K$, is $\tfrac{1}{2} + \tfrac{1}{2} \sqrt{1- 8\sigma^2/K^2}$.  Note this is independent of $r$, and increases with $\sigma/K$, the larger the noise $\sigma_g$ as a fraction of the carrying capacity $K$.  Population sizes can also be inflated relative to their deterministic equilibrium whenever the second derivative is positive (as in the lower equilibrium in the May model of alternative stable states, considered below.)  As predicted by the system size expansion, the deviation from the deterministic result is of order $N^{-1}$.

@@

# Parameters
p <- list(r = .4, K = 1, sigma = .15, N = 500)
equib <- (1 + sqrt(1 - 8 * p$sigma^2/p$K^2) )/ 2

# BUGS model
logistic  <- nimble::nimbleCode({
  x[1] <- x0
  for(t in 1:(N-1)){
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)
    x[t+1] ~ dnorm(mu[t], sd = sigma)
  }
})

# Compile and run model
model <- 
nimbleModel(logistic,
            constants = p,
            inits = list(x0 = 1))
cmodel <- compileNimble(model)
set.seed(123)

# Perform the simulations
df <- map_dfr(1:10000, 
              function(rep){
                simulate(cmodel)
                data.frame(t = seq_along(cmodel$x), x = cmodel$x)
              },
              .id = "rep")

@@

Note that the average across stochastic replicates is slightly less than K=1 (black line),
and matches almost perfectly the equilibrium predicted by the higher order correction theory.
The standard deviation of fluctuations shown for reference as well.

```{r}
read_csv("inflation.csv") %>% 
  gather(statistic, value, -t) %>%
  ggplot(aes(t,value, col=statistic, lty=statistic)) + 
  geom_line(lwd=1) + 
  geom_hline(yintercept = 1, col="black") +
  scale_color_ptol() + 
  coord_cartesian(xlim=c(100,500), ylim=c(0.1, 1.2))
```

### 2018-03-06 paper/paper.Rmd
tweaking refs

abstract: |
  # Article Type: Review & Synthesis
  # Running title: Noise phenomena
  # Figures: 4
  # Tables: 0
  # Word count: 
  # Word count in abstract: 
  # Number of References: 
  # Archive statement:

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = FALSE)
library(tidyverse)
library(gridExtra)
library(ggthemes)
library(png)
colours <- ptol_pal()(2)

#https://stackoverflow.com/a/24241954/258662
fancy_scientific <- function(l) { 
     l <- format(l, scientific = TRUE)
     l <- gsub("^(.*)e", "'\\1'e", l)
     l <- gsub("e", "%*%10^", l)
     l <- gsub("^'1'%\\*%", "", l)
     l <- gsub("\\+", "", l)
     parse(text=l)
}
```

@@

<!-- -->

@@

Since @Nisbet1976, examples of stochastic resonance have been demonstrated in common
models and implicated in empirical examples of increasing complexity, [@Greenman2003],
particularly in higher-dimensional disease and outbreak models [@Blarer1999; @Greeman2003; @McKane2005; @Alonso2007; @Black2010].
@Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod.  @McKane2005, @Alonso2007, and @Black2010 illustrate
the emergence of quasi-cycles in common disease outbreak models, including
the use of power spectrum analysis.  
@Pineda-Krch2007 compare patterns from stochastically generated cycles to
stable limit cycles they resemble to ask how we can discriminate between
these mechanisms.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern)
[e.g. @Gardiner2009; @Black2010].

@@

<!--Figure 3: Noisy switch -->
```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Panel A: Data on prevelance of rubella in Copenhagen illustrating potential stochastic switching dynamics between two regimes, reproduced from @Keeling2001. B-D. Stochastic switiching in May's model of alternative stable state dynamics. Panel B: timeseries showning stochastic switching events between regimes. Panel C: separate nonlinear birth and death processes intersect at multiple points creating alternate stable states. Panel D: The corresponding potential well derived from the birth-death model: $U(x) = -\\int b(x) - d(x) \\ ud x $. Parameter values shown r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000, see code in Appendix A. \\label{noisy_switch}"}

@@

pop_dai <- ggplot() + 
  annotation_raster(readPNG("figures/pop-dai2015.png"), 
                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
  labs(subtitle = "A.")
cv_dai <- ggplot() + 
  annotation_raster(readPNG("figures/cv-dai2015.png"), 
                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
  labs(subtitle = "B.")
pop_sim <- 
  tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
  scale_y_log10(labels=fancy_scientific)  + 
  coord_cartesian(ylim=c(1e3, 5e5)) +
  labs(subtitle = "C. Yeast population size", 
       caption = "Vertical red dashed line indicates tipping point location")

@@

More broadly, examples of early warning signals such as increasing variance,
auto-correlation, and flickering all point to ways in which stochasticity
and noise-driven phenomena can be transformed from something which obscures
underlying processes to something which reveals them.  I believe this is
role could apply far more broadly than in just the context of stability and
tipping points.  The examples from fold bifurcations already suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].

Interpreting stochastic processes as potential new sources of information
has also proven to be a
rich catalyst for bringing these concepts beyond the realm of abstract
theory. A rich array of experimental and observational tests of early
warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017;
and many others, _e.g._ see @Scheffer2015 review] has served not merely
to illustrate these phenomena can be seen in laboratory and field ecosystems,
but also to spark suggestions of additional possible indicators and surface
a wide array of practical [@Boettiger2013b; @Scheffer2015], statistical
[e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015;
@Hastings2010] challenges to such application.  This creates a two-way dialog
between empirical and theoretical work that does much to advance the field
as a whole.

<!--
One of the most interesting aspects of noise is as a source of
additional information.  This can be realized in a variety of ways.
For instance, in deriving expressions for the variance created by
demographic noise above, we observed that different partitions of
a logistic growth model can lead to the same macroscopic dynamics
(same values for $b$ - $d$), while having different variance (due
to different values for $b$+$d$).  This suggests that the variation
observed in such a population could potentially help resolve
how dynamics are partitioned between individual birth and death
rates (though in practice one would have to first account for any
contribution of environmental noise).
-->

@@

>>>>

abstract: |
  # Article Type: Review & Synthesis
  # Running title: Noise phenomena
  # Figures: 4
  # Tables: 0
  # Word count: 
  # Word count in abstract: 
  # Number of References: 117
  # Archive statement:

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = FALSE)
library(tidyverse)
library(gridExtra)
library(ggthemes)
library(png)
library(scales)
colours <- ptol_pal()(2)

#https://stackoverflow.com/a/24241954/258662
scientific <- function(l) {
    if(max(l, na.rm=TRUE) < 1e3) return(l)
     l <- format(l, scientific = TRUE)
     l <- gsub("^(.*)e", "'\\1'e", l)
     l <- gsub("e", "%*%10^", l)
     l <- gsub("^'1'%\\*%", "", l)
     l <- gsub("\\+", "", l)
     parse(text=l)
}
```

@@

@@

Since @Nisbet1976, examples of stochastic resonance have been demonstrated in common
models and implicated in empirical examples of increasing complexity, [@Greenman2003],
particularly in higher-dimensional disease and outbreak models [@Blarer1999; @Greenman2003; @McKane2005; @Alonso2007; @Black2010].
@Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod.  @McKane2005, @Alonso2007, and @Black2010 illustrate
the emergence of quasi-cycles in common disease outbreak models, including
the use of power spectrum analysis.  
@Pineda-Krch2007 compare patterns from stochastically generated cycles to
stable limit cycles they resemble to ask how we can discriminate between
these mechanisms.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern)
[e.g. @Gardiner2009; @Black2010].

@@

<!--Figure 3: Noisy switch -->
```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Panel A: Data on prevelance of rubella in Copenhagen illustrating potential stochastic switching dynamics between two regimes, reproduced from @Keeling2001. B-D. Stochastic switiching in May's model of alternative stable state dynamics. Panel B: timeseries showning stochastic switching events between regimes. Panel C: separate nonlinear birth and death processes intersect at multiple points creating alternate stable states. Panel D: The corresponding potential well derived from the birth-death model: $U(x) = -\\int b(x) - d(x) \\ud x$. Parameter values shown r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000, see code in Appendix A. \\label{noisy_switch}"}

@@

pop_dai <- ggplot() + 
  annotation_raster(readPNG("figures/pop-dai2015.png"), 
                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
  labs(subtitle = "A.")
cv_dai <- ggplot() + 
  annotation_raster(readPNG("figures/cv-dai2015.png"), 
                    xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
  labs(subtitle = "B.")
pop_sim <- 
  tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line(col=colours[1]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[2], lty=2) + 
  scale_y_log10(labels=scientific)  + 
  coord_cartesian(ylim=c(1e3, 5e5)) +
  labs(subtitle = "C. Yeast population size", 
       caption = "Vertical red dashed line indicates tipping point location")

@@

Interpreting stochastic processes as potential new sources of information
has also proven to be an effective catalyst in bringing concepts from theory 
to application. An array of experimental and observational tests of early
warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017; @Scheffer2015] 
has illustrated where these phenomena can be seen in laboratory and field ecosystems,
and also sparked discussion of
a wide array of practical [@Boettiger2013b; @Scheffer2015], statistical
[e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015;
@Hastings2010] challenges that arise.  This creates a two-way dialog
between empirical and theoretical work that does much to advance the field
as a whole. Similar dialog between theory and experiment in other areas
of stochastic dynamics could benefit both sides.

The potential for stochastic phenomena to provide information about underlying
processes goes well beyond avialable examples in stability and early warning signals.
The examples from fold bifurcations already suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].  Noise can provide
information in unrelated ways as well. in deriving expressions for the variance created by
demographic noise above, we observed that different partitions of
a logistic growth model can lead to the same macroscopic dynamics
(same values for $b$ - $d$), while having different variance (due
to different values for $b$+$d$).  This suggests that the variation
observed in such a population could potentially help resolve
how dynamics are partitioned between individual birth and death
rates (though in practice one would have to first account for any
contribution of environmental noise).

@@

# Acknowledgements

The author acknowledges feedback and advice from the editor, 
Tim Coulson and two anonymous reviewers. This work was supported in
part by  USDA National Institute of Food and Agriculture, Hatch
project CA-B-INS-0162-H.
